---
title: "OSM Working Draft"
author: "Eugene Chong"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: 
      collapsed: false
    toc_depth: 4
    code_folding: show
---
  
```{r setup, include=FALSE}
cache_state = TRUE

knitr::opts_chunk$set(echo = TRUE, cache = cache_state, message = FALSE)
```

# Admin

Load packages

```{r packages}
library(tidyverse)
library(sf)
library(osmdata)
library(ggmap)

library(httr)
library(jsonlite)

# unload packages
# detach("package:microbenchmark", unload=TRUE)
```

# Explore OSM data 

**Resources**

Vignette: https://cran.r-project.org/web/packages/osmdata/vignettes/osmdata.html

Map features: https://wiki.openstreetmap.org/wiki/Map_Features

`?available_features()` and `?available_tags([feature])`

## Baltimore

### Query parameters

List of parameters:

* City name
* key
* value

```{r}
Balt_bbox <- getbb("Baltimore") %>% 
  opq()
```

### Transit

Query the bus stops in Baltimore.

```{r}
Balt_busStops <- Balt_bbox %>% 
  add_osm_feature(key = "highway", value = "bus_stop") %>% 
  osmdata_sf() %>% 
  # keep only the points. Note that the query returned 4333 points, 9 polygons, and 1 multi-line feature
  .$osm_points
```

What's in the bus stops data?

```{r}
glimpse(Balt_busStops)
```

Make a quick map

```{r}
Balt_map <- get_map(getbb("Baltimore"), maptype = "toner-background")
```

```{r, fig.height=10, fig.width=7}
ggmap(Balt_map) +
  geom_sf(data = Balt_busStops,
          inherit.aes = FALSE, # this is necessary
          alpha = 0.5) +
  labs(title = "Bus Stops in Baltimore",
       caption = "Data source: OSM",
       x = "",
       y = "")
```

### Query some historical data

#### API admin

```{r}
base_url <- "https://api.ohsome.org/v0.9"
api_metadata <- GET(url= paste(base_url, "/metadata", sep = "")) %>% 
  content("text") %>% 
  fromJSON()
```

The below tells us the database contains data from October 8, 2007 to May 23, 2020.

```{r}
api_metadata$extractRegion$temporalExtent
```

#### Aggregation endpoint

Use this endpoint to aggregate OSM data, e.g. counts, areas, lengths, and users (contributors).

Let's look at the trends for mapping bus stops in Baltimore.

```{r}

# api_bbox <- getbb("Baltimore") %>% bbox_to_string() # note that the order of the coords is flipped from what the database needs
balt_bbox <- "-76.770759, 39.1308816,-76.450759,39.4508816"
# api_bbox <- "8.6581,49.3836,8.7225,49.4363"

api_keys <- "highway"
api_values <- "bus_stop"

monthly <- "2007-11-01/2020-05-23/P1M"

api_data <- GET(url = paste(base_url, "/elements/count", sep = ""),
                query = list(
                  bboxes = balt_bbox,
                  keys = api_keys,
                  values = api_values,
                  time = monthly))

busStops_hist <- content(api_data, as = "text") %>% 
  fromJSON() %>% 
  .$result
```

The query from `osmdata` showed 4333 bus stops in Baltimore currently. The OSHDB query shows 4241 bus stops as of May 1, 2020.

```{r}
ggplot(busStops_hist, 
       aes(x = as.Date(timestamp),
           y = value)) +
  geom_line() +
  theme_bw() +
  labs(title = "Bus Stops in Baltimore Mapped In OSM Over Time",
       caption = "Source: OSHDB, ohsome API",
       x = "Date",
       y = "Count")
```

#### Extraction endpoint

Use this endpoint to pull historical snapshots of OSM features.

**Code below is not the most elegant and the resulting dataframe seems odd - is there a better way to convert the API geoJSON response into sf?**

```{r}
# is there a more elegant way to do the below?
file <- tempfile(fileext = ".geojson")
extraction_api_data <- GET(url= paste(base_url, "/elements/geometry", sep = ""),
                query = list(
                  bboxes = balt_bbox,
                  keys = api_keys,
                  values = api_values,
                  types = "point", # do I want to limit it to points only?
                  time = monthly),
                httr::write_disk(file))
busStops_geom_hist <- sf::read_sf(file)
busStops_geom_hist$X.snapshotTimestamp <- as.Date(busStops_geom_hist$X.snapshotTimestamp)

busStops_geom_hist

# busStops_geom_hist <- content(extraction_api_data, as = "text") %>%
#   fromJSON()

# write(busStops_geom_hist %>% toJSON(), "test.json")
```

According to our aggregated data, there was a massive spike in bus stops in Baltimore on OSM at the beginning of February, 2019: from 280 to 3968.

First, does the geometry data have the same number of observations as the aggregated data? It's very close - it may be a matter of the time of day queried.

```{r}
busStops_changeMap <- busStops_geom_hist %>% 
  filter(X.snapshotTimestamp %in% as.Date(c("2019-02-01", "2019-01-01")))

busStops_changeMap %>% group_by(X.snapshotTimestamp) %>% 
  summarize(count = n())
```

Next, compare to two months side-by-side on a map. 

```{r, fig.width=12, fig.height=8}
ggmap(Balt_map) +
  geom_sf(data = busStops_changeMap,
          inherit.aes = FALSE, 
          alpha = 0.5) +
  labs(title = "Change in Bus Stops Mapped on OSM in Baltimore",
       subtitle = "Number of bus stops jumped from 280 in January 2019 to 3968 in February 2019.",
       caption = "Data source: OSHDB",
       x = "",
       y = "") +
  facet_wrap(~ X.snapshotTimestamp) 
```

